---
title: "ETC5513 Assignment 3"
author: "Mey Siv and Devaansh Gupta"
editor: visual
format: 
  html: 
    theme: cerulean
  pdf: default
execute:
    echo: true
    warning: false
    message: false
---
```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(kableExtra)
library(tidyr)
```

## Methodology

```{r data-cleaning, message=FALSE, warning=FALSE}
# Read the energy data - using relative path for team compatibility
energy <- read.csv("owid-energy-data.csv")

countries <- c("Australia", "India", "United States", "China", "United Kingdom")
years <- 2019:2023

selected_vars <- c(
  "country", "year", "population", "gdp", "biofuel_consumption", "biofuel_cons_per_capita",
  "coal_consumption", "coal_cons_per_capita", "energy_cons_change_twh", "energy_per_capita",
  "energy_per_gdp", "fossil_cons_change_twh", "fossil_elec_per_capita", "fossil_energy_per_capita",
  "fossil_fuel_consumption", "gas_consumption", "gas_elec_per_capita", "gas_energy_per_capita",
  "hydro_consumption", "hydro_energy_per_capita", "low_carbon_consumption",
  "low_carbon_energy_per_capita", "nuclear_consumption", "nuclear_energy_per_capita",
  "oil_consumption", "oil_energy_per_capita", "other_renewable_consumption",
  "other_renewables_energy_per_capita", "primary_energy_consumption",
  "renewables_consumption", "renewables_energy_per_capita", "solar_consumption",
  "solar_energy_per_capita", "wind_consumption", "wind_energy_per_capita"
)

energy_filtered <- energy %>%
  filter(country %in% countries, year %in% years) %>%
  select(all_of(selected_vars)) %>%
  drop_na(population)
```

Five major economies—Australia, India, the United States, China, and the United Kingdom—are the focus of this analysis, which makes use of the "Our World in Data" global energy dataset. These nations were chosen due to their varied economic situations, population sizes, and energy regulations. The study, which spans 2019–2023, attempts to evaluate how patterns of energy consumption—specifically, fossil versus renewable sources—have changed recently.

Only pertinent energy metrics, including population, GDP, total and per capita use of fossil fuels and renewables, and other low-carbon sources including hydro, wind, and solar, were kept in a filtered subset of the dataset. For important variables like population, missing data were eliminated to guarantee consistency. Both absolute and intensity-based energy use were reflected in the selection of key variables.

Each country's average energy consumption over the course of five years is shown in the table, along with the average energy consumption from fossil fuels and renewables per capita. We may assess the degree of integration of renewable energy in national energy profiles using this comparative table.

Facet plots are used in the figure to display time-series trends of both fossil and renewable energy consumption, allowing for a side-by-side comparison of changes in energy patterns across nations. This graphic aids in determining if nations are shifting away from fossil fuels and toward clean energy.

To ensure complete reproducibility, all analyses and visualizations were carried out in R utilizing tidyverse tools through R Markdown. The use of chunk options (`echo=FALSE`, `message=FALSE`, `warning=FALSE`) ensures that code does not distract from the professional presentation of the report, while still enabling seamless re-execution if the data or methods are updated.

### Table 1: Summary of average energy metrics (2019–2023) for selected countries.

```{r summary-table, echo=FALSE, results='asis'}
summary_table <- energy_filtered %>%
  group_by(country) %>%
  summarise(
    `Avg. Renewable Per Capita (kWh)` = round(mean(renewables_energy_per_capita, na.rm = TRUE), 1),
    `Avg. Fossil Per Capita (kWh)` = round(mean(fossil_energy_per_capita, na.rm = TRUE), 1),
    `Renewable Share (%)` = round(mean(renewables_consumption / primary_energy_consumption * 100, na.rm = TRUE), 1)
  )

kable(summary_table, caption = "Table 1: Summary of average energy metrics (2019–2023) for selected countries.") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```	


{r facet-trends, echo=FALSE, fig.cap="Fossil vs. renewable energy consumption (TWh) in each country, 2019–2023.", fig.width=10, fig.height=6}
library(scales)
library(tidyr)


energy_long <- energy_filtered %>%
  select(country, year, fossil_fuel_consumption, renewables_consumption) %>%
  pivot_longer(cols = c(fossil_fuel_consumption, renewables_consumption),
               names_to = "type", values_to = "consumption") %>%
  mutate(type = ifelse(type == "fossil_fuel_consumption", "Fossil", "Renewables"))

ggplot(energy_long, aes(x = year, y = consumption, color = type)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  facet_wrap(~country, scales = "free_y") +
  labs(x = "Year", y = "Energy Consumption (TWh)", color = "Type") +
  theme_minimal() +
  theme(legend.position = "top")


