---
title: "ETC5513 Assignment 3"
author: "Mey Siv and Devaansh Gupta"
editor: visual
format:
  html:
    theme: cerulean
    toc: true
    toc-depth: 2
  pdf: 
    toc: true
    toc-depth: 2
    number-sections: true
execute:
    echo: false
    warning: false
    message: false
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(kableExtra)
library(tidyr)
```

## Executive Summary

Australia, India, the United States, China, and the United Kingdom's worldwide energy consumption patterns from 2019 to 2023 are examined in this paper. It examines how low-carbon and renewable energy sources are replacing fossil fuels. Important variables are looked at, including GDP, population, and several energy measurements. The results provide credence to current debates about the economic ramifications and sustainability of energy.

## Introduction

Energy use is essential for both environmental sustainability and economic growth. Countries have been focusing more and more on low-carbon and renewable energy sources in recent years. Australia, India, the United States, China, and the United Kingdom are the five major economies whose energy consumption trends are evaluated in this paper from 2019 to 2023. These nations were chosen because of their varied economic systems, population sizes, and energy strategies. Understanding the evolution of energy demand, particularly from fossil and renewable sources, is the goal. We also take into account the correlation between these changes and changes in the population and GDP. The goal of the analysis is to highlight both domestic and international trends related to energy policy. To guarantee clarity and reproducibility, R and R Markdown are used for all analyses.

## Methodology

```{r data-cleaning, message=FALSE, warning=FALSE}

energy <- read.csv("owid-energy-data.csv")

countries <- c("Australia", "India", "United States", "China", "United Kingdom")
years <- 2019:2023

selected_vars <- c(
  "country", "year", "population", "gdp", "biofuel_consumption", "biofuel_cons_per_capita",
  "coal_consumption", "coal_cons_per_capita", "energy_cons_change_twh", "energy_per_capita",
  "energy_per_gdp", "fossil_cons_change_twh", "fossil_elec_per_capita", "fossil_energy_per_capita",
  "fossil_fuel_consumption", "gas_consumption", "gas_elec_per_capita", "gas_energy_per_capita",
  "hydro_consumption", "hydro_energy_per_capita", "low_carbon_consumption",
  "low_carbon_energy_per_capita", "nuclear_consumption", "nuclear_energy_per_capita",
  "oil_consumption", "oil_energy_per_capita", "other_renewable_consumption",
  "other_renewables_energy_per_capita", "primary_energy_consumption",
  "renewables_consumption", "renewables_energy_per_capita", "solar_consumption",
  "solar_energy_per_capita", "wind_consumption", "wind_energy_per_capita"
)

energy_filtered <- energy %>%
  filter(country %in% countries, year %in% years) %>%
  select(all_of(selected_vars)) %>%
  drop_na(population)
```

Five major economies—Australia, India, the United States, China, and the United Kingdom—are the focus of this analysis, which makes use of the "Our World in Data" global energy dataset. These nations were chosen due to their varied economic situations, population sizes, and energy regulations. The study, which spans 2019–2023, attempts to evaluate how patterns of energy consumption—specifically, fossil versus renewable sources—have changed recently.

Only pertinent energy metrics, including population, GDP, total and per capita use of fossil fuels and renewables, and other low-carbon sources including hydro, wind, and solar, were kept in a filtered subset of the dataset. For important variables like population, missing data were eliminated to guarantee consistency. Both absolute and intensity-based energy use were reflected in the selection of key variables.

Each country's average energy consumption over the course of five years is shown in the table, along with the average energy consumption from fossil fuels and renewables per capita. We may assess the degree of integration of renewable energy in national energy profiles using this comparative table.

Facet plots are used in the figure to display time-series trends of both fossil and renewable energy consumption, allowing for a side-by-side comparison of changes in energy patterns across nations. This graphic aids in determining if nations are shifting away from fossil fuels and toward clean energy.

To ensure complete reproducibility, all analyses and visualizations were carried out in R utilizing tidyverse tools through R Markdown. The use of chunk options (`echo=FALSE`, `message=FALSE`, `warning=FALSE`) ensures that code does not distract from the professional presentation of the report, while still enabling seamless re-execution if the data or methods are updated.

Table 1: Summary of average energy metrics (2019–2023) for selected countries.

```{r summary-table, echo=FALSE, results='asis'}
#| label: tbl-summary
#| tbl-cap: "Summary of average energy metrics (2019–2023)"


summary_table <- energy_filtered %>%
  group_by(country) %>%
  summarise(
    `Avg. Renewable Per Capita (kWh)` = round(mean(renewables_energy_per_capita, na.rm = TRUE), 1),
    `Avg. Fossil Per Capita (kWh)` = round(mean(fossil_energy_per_capita, na.rm = TRUE), 1),
    `Renewable Share (%)` = round(mean(renewables_consumption / primary_energy_consumption * 100, na.rm = TRUE), 1)
  )


knitr::kable(summary_table) %>%
    kable_styling(
    latex_options = c("hold_position", "repeat_header"),  
    font_size = 10)                                        

```

```{r facet-trends, echo=FALSE, fig.cap="...", fig.width=10, fig.height=6}

library(scales) 
library(tidyr)

energy_long <- energy_filtered %>% select(country, year, fossil_fuel_consumption, renewables_consumption) %>% pivot_longer(cols = c(fossil_fuel_consumption, renewables_consumption), names_to = "type", values_to = "consumption") %>% mutate(type = ifelse(type == "fossil_fuel_consumption", "Fossil", "Renewables"))

ggplot(energy_long, aes(x = year, y = consumption, color = type)) + geom_line(linewidth = 1.2) + geom_point(size = 2) + facet_wrap(~country, scales = "free_y") + labs(x = "Year", y = "Energy Consumption (TWh)", color = "Type") + theme_minimal() + theme(legend.position = "top")

```

## Result

```{r, echo=FALSE}
#| label: fig-renewable
#| fig-cap: "Renewable Energy Consumption by Source (2019–2023)"
renewables_compare <- energy_filtered %>%
  filter(country %in% c("Australia", "India", "United States", "China", "United Kingdom")) %>%
  select(country, year, solar_consumption, wind_consumption, hydro_consumption,
         other_renewable_consumption, biofuel_consumption) %>%
  pivot_longer(cols = -c(country, year), names_to = "source", values_to = "consumption") %>%
  mutate(source = case_when(
    source == "solar_consumption" ~ "Solar",
    source == "wind_consumption" ~ "Wind",
    source == "hydro_consumption" ~ "Hydro",
    source == "other_renewable_consumption" ~ "Other Renewables",
    source == "biofuel_consumption" ~ "Biofuel"
  ))

ggplot(renewables_compare, aes(x = factor(year), y = consumption, fill = source)) +
  geom_col(position = "dodge") +
  facet_wrap(~ country, scales = "free_y") +
  labs(
    title = "Renewable Energy Consumption by Source (2019–2023)",
    subtitle = "Comparison Across Australia, India, United States, China, and United Kingdom",
    x = "Year",
    y = "Consumption (TWh)",
    fill = "Energy Source"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

@fig-renewable shows total renewable energy consumption. Australia has significantly adopted solar power due to government incentives and subsidies for rooftop solar installations. In contrast, the United States and the United Kingdom rely more on wind energy, given their larger land areas and favorable wind conditions. Meanwhile, China and India have invested heavily in hydropower, taking advantage of vast river systems, high rainfall, and significant elevation changes.

```{r}
#| label: fig-fossil
#| echo: false
#| fig-cap: "Fossil Fuel Energy Consumption by Source (2019–2023)"
fossilfuels_compare <- energy_filtered %>%
  filter(country %in% c("Australia", "India", "United States", "China", "United Kingdom")) %>%
  select(country, year, coal_consumption, oil_consumption, gas_consumption) %>%
  pivot_longer(cols = -c(country, year), names_to = "source", values_to = "consumption") %>%
  mutate(source = case_when(
    source == "coal_consumption" ~ "Coal",
    source == "oil_consumption" ~ "Oil",
    source == "gas_consumption" ~ "Gas"
  ))

# Plot the fossil fuel consumption
ggplot(fossilfuels_compare, aes(x = factor(year), y = consumption, fill = source)) +
  geom_col(position = "dodge") +
  facet_wrap(~ country, scales = "free_y") +
  labs(
    title = "Fossil Fuel Consumption by Source (2019–2023)",
    subtitle = "Comparison Across Australia, India, United States, China, and United Kingdom",
    x = "Year",
    y = "Consumption (TWh)",
    fill = "Fossil Fuel Source"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

@fig-fossil illustrates total fossil fuel consumption. China and India rely heavily on coal for energy due to its domestic affordability and widespread use in power generation. Conversely, Australia, the UK, and the US primarily consume oil and gas for heating, power generation, and transport. Because of its large population—second only to India—China depends significantly on both fossil fuels and renewables to meet its energy demands.

@tbl-summary shows per capita energy consumption. While these countries are gradually shifting towards renewables, fossil fuels still dominate their overall energy mix. China’s per capita energy use is lower than that of Australia and the United States, which lead in both fossil fuel and renewable energy consumption. Differences in population size, economic structure, geography, and lifestyle drive these variations.


## Conclusion

This analysis examined energy consumption trends (2019–2023) in Australia, India, the United States, China, and the United Kingdom, focusing on the shift from fossil fuels to low-carbon and renewable sources. Key factors such as GDP, population, and per capita energy use were evaluated to assess sustainability and economic implications.

Fossil fuels remain dominant across all five economies, but renewable energy adoption varies significantly. China leads in total consumption due to its large population and industrial demand, though per capita usage remains moderate. The UK has the highest renewable share (\~19%), supported by strong policy incentives, while India’s renewable growth lags despite rising energy needs. Australia and the US exhibit high per capita consumption, reflecting energy-intensive economies and lifestyles.

These trends underscore the influence of economic structure, resource availability, and policy frameworks on national energy strategies. While progress in renewables is evident, fossil fuels still underpin global energy systems, highlighting the need for accelerated transitions to meet sustainability goals.

## Recommendation

To accelerate the transition to sustainable energy, the following actions should be prioritized:

-   Expand Renewable Energy Deployment – Fast-track solar, wind, and other renewable projects to reduce reliance on fossil fuels.

-   Modernize Grid Infrastructure – Upgrade power grids with smart technologies and large-scale battery storage to manage renewable energy variability.

-   Strengthen Policy Support – Implement carbon pricing (taxes or cap-and-trade systems) and enhance incentives for clean energy research, development, and adoption.

## Reference

Hannah Ritchie, Pablo Rosado, and Max Roser (2023) - “Energy” Published online at OurWorldinData.org. Retrieved from: 'https://ourworldindata.org/energy' \[Online Resource\]
