---
title: "Global Energy Consumption Trends (2019–2023)"
author: 
  - Devaansh
  - Mey
format: 
  revealjs:
    theme: solarized
    slide-number: true
    toc: false
    transition: slide
    self-contained: true
execute: 
  echo: false
  message: false
  warning: false
---


## Title Slide

### Global Energy Consumption Trends (2019–2023)

*Presented by:* Devaansh, Mey\
ETC5513 - Group Presentation\
Week 12 Tutorial

---

## Topics to be covered for today
- Background to the global energy transitions
- Dataset Description
- Methodology
- Results
- Conclusion
- Recommendation

---

## Problem Introduction

-   Energy systems are central to global economic activity and climate strategy.
-   Nations must transition from fossil fuels to renewables for long-term sustainability.
-   This study examines the energy profiles of five key countries:\
    Australia, India, United States, China, and United Kingdom.

------------------------------------------------------------------------

## Dataset Description

-   *Source:* Our World in Data – Global Energy Dataset
-   *Years covered:* 2019 to 2023
-   *Countries selected:* Based on energy diversity and geopolitical relevance
-   *Key variables:*
    -   Countries, GDP, Year
    -   Fossil and renewable energy consumption (total and per capita)

------------------------------------------------------------------------

## Methodology

-   Filtered dataset to isolate relevant years and countries
-   Selected energy metrics related to fossil and renewable use
-   Computed averages and trends using dplyr and tidyr
-   Visualized temporal patterns using ggplot2
-   Full reproducibility ensured via R and Quarto integration

------------------------------------------------------------------------



```{r data-cleaning, message=FALSE, warning=FALSE, echo=FALSE}

library(dplyr)
library(tidyr)
library(ggplot2)


energy <- read.csv("data/owid-energy-data.csv")
countries <- c("Australia", "India", "United States", "China", "United Kingdom")
years <- 2019:2023

selected_vars <- c(
  "country", "year", "population", "gdp", "biofuel_consumption", "biofuel_cons_per_capita",
  "coal_consumption", "coal_cons_per_capita", "energy_cons_change_twh", "energy_per_capita",
  "energy_per_gdp", "fossil_cons_change_twh", "fossil_elec_per_capita", "fossil_energy_per_capita",
  "fossil_fuel_consumption", "gas_consumption", "gas_elec_per_capita", "gas_energy_per_capita",
  "hydro_consumption", "hydro_energy_per_capita", "low_carbon_consumption",
  "low_carbon_energy_per_capita", "nuclear_consumption", "nuclear_energy_per_capita",
  "oil_consumption", "oil_energy_per_capita", "other_renewable_consumption",
  "other_renewables_energy_per_capita", "primary_energy_consumption",
  "renewables_consumption", "renewables_energy_per_capita", "solar_consumption",
  "solar_energy_per_capita", "wind_consumption", "wind_energy_per_capita"
)

energy_filtered <- energy %>%
  filter(country %in% countries, year %in% years) %>%
  select(all_of(selected_vars)) %>%
  drop_na(population)
```
---

## Countries with the renewable energy

```{r }
# Filter and reshape the data
renewables_compare <- energy_filtered %>%
  filter(country %in% c("Australia", "India", "United States", "China", "United Kingdom")) %>%
  select(country, year, solar_consumption, wind_consumption, hydro_consumption,
         other_renewable_consumption, biofuel_consumption) %>%
  pivot_longer(cols = -c(country, year), names_to = "source", values_to = "consumption") %>%
  mutate(source = case_when(
    source == "solar_consumption" ~ "Solar",
    source == "wind_consumption" ~ "Wind",
    source == "hydro_consumption" ~ "Hydro",
    source == "other_renewable_consumption" ~ "Other Renewables",
    source == "biofuel_consumption" ~ "Biofuel"
  ))


ggplot(renewables_compare, aes(x = factor(year), y = consumption, fill = source)) +
  geom_col(position = "dodge") +
  facet_wrap(~ country, scales = "free_y") +
  labs(
    title = "Renewable Energy Consumption by Source (2019–2023)",
    subtitle = "Comparison Across Australia, India, United States, China, and United Kingdom",
    x = "Year",
    y = "Consumption (TWh)",
    fill = "Energy Source"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

---
## Countries with the Fossil Fuel Consumption by Source

```{r}
# Filter and reshape the data for fossil fuels
fossilfuels_compare <- energy_filtered %>%
  filter(country %in% c("Australia", "India", "United States", "China", "United Kingdom")) %>%
  select(country, year, coal_consumption, oil_consumption, gas_consumption) %>%
  pivot_longer(cols = -c(country, year), names_to = "source", values_to = "consumption") %>%
  mutate(source = case_when(
    source == "coal_consumption" ~ "Coal",
    source == "oil_consumption" ~ "Oil",
    source == "gas_consumption" ~ "Gas"
  ))

# Plot the fossil fuel consumption
ggplot(fossilfuels_compare, aes(x = factor(year), y = consumption, fill = source)) +
  geom_col(position = "dodge") +
  facet_wrap(~ country, scales = "free_y") +
  labs(
    title = "Fossil Fuel Consumption by Source (2019–2023)",
    subtitle = "Comparison Across Australia, India, United States, China, and United Kingdom",
    x = "Year",
    y = "Consumption (TWh)",
    fill = "Fossil Fuel Source"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```
---

## Summary of average energy metrics (2019–2023)

```{r summary-table, echo=FALSE, results='asis'}
#| label: tbl-summary
#| tbl-cap: "Summary of average energy metrics (2019–2023)"
library(kableExtra)

summary_table <- energy_filtered %>%
  group_by(country) %>%
  summarise(
    `Avg. Renewable Per Capita (kWh)` = round(mean(renewables_energy_per_capita, na.rm = TRUE), 1),
    `Avg. Fossil Per Capita (kWh)` = round(mean(fossil_energy_per_capita, na.rm = TRUE), 1),
    `Renewable Share (%)` = round(mean(renewables_consumption / primary_energy_consumption * 100, na.rm = TRUE), 1)
  )


knitr::kable(summary_table) %>%
    kable_styling(
    latex_options = c("hold_position", "repeat_header"))

```

---

## Conslusion

-   Fossil fuel use remains dominant in all five countries, though renewables are growing.

-   China uses both fossil fuels and renewables heavily, but per capita energy use is lower due to its large population.

-   The UK has the highest renewable share (\~19%), driven by wind and policy support.

-   India’s renewable adoption lags behind despite high demand, but hydropower plays a significant role.

-   Australia and the US have higher per capita energy use, reflecting energy-intensive industries and high living standards.

-   Choices are influences by Economic structure, resources, and policy choices shape each country’s energy mix.

---

## Recommendation :

To accelerate the transition to sustainable energy, the following actions should be prioritized:

-  Expand Renewable Energy Deployment 

- Modernize Grid Infrastructure 

- Strengthen Policy Support 